\documentclass[pdflatex,sn-mathphys-ay]{sn-jnl}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}

\begin{document}

\title[GEE-LLM: Grounded GEE Code Gen and Scientific Interpretation]{GEE-LLM: Grounded Google Earth Engine Code Generation and Scientific Interpretation for Geospatial Decision Support}

\author[1]{\fnm{Aman} \sur{Sharma}}
\author*[1,2]{\fnm{Manish} \sur{Kumar}}\email{naik.manish17@gmail.com}

\affil[1]{\orgname{Centre for Green Energy and Nanotechnology (C-GENT), Himachal Pradesh University (HPU)}, \orgaddress{\city{Shimla}, \state{Himachal Pradesh}, \postcode{171005}, \country{India}}}
\affil[2]{\orgname{University Institute of Technology (UIT), Himachal Pradesh University (HPU)}, \orgaddress{\city{Shimla}, \state{Himachal Pradesh}, \postcode{171005}, \country{India}}}

\maketitle

\section*{Highlights}
\begin{itemize}
    \item GEE-LLM closes the complete loop from natural language query to decision-ready geospatial intelligence, combining verified satellite code generation with grounded scientific interpretation.
    \item A two-pass large language model architecture separates executable code generation from result interpretation, ensuring that the reasoning stage operates exclusively on execution-validated numeric output.
    \item A local sandboxed execution engine with physical boundary validation raises script correctness from 12--17\% (zero-shot baseline) to 44--45\% across a 100-query benchmark spanning diverse Indian ecological zones.
    \item McNemar's test confirms that improvements are statistically significant ($p < 0.001$), with odds ratios up to 15.0, demonstrating that the self-correction loop recovers failing scripts rather than degrading successful ones.
    \item The framework operates on commodity CPU hardware without GPU resources or commercial cloud subscriptions, making geospatial decision support accessible to resource-constrained research environments.
\end{itemize}

\section*{Statements and Declarations}

\subsection*{Funding}
No funds, grants, or other support was received.

\subsection*{Competing Interests}
The authors have no relevant financial or non-financial interests to disclose.

\subsection*{Ethics Approval}
Not applicable.

\subsection*{Consent to Participate}
Not applicable.

\subsection*{Consent for Publication}
Not applicable.

\subsection*{Data and Code Availability}
The interactive web application is hosted at: \url{https://geo-ai-llm-aman.streamlit.app/}. The full GEE-LLM software framework, the lexical template corpus, the 100-query benchmark, and all model output logs are publicly available at: \url{https://github.com/AmanSharma000/gee-llm}.

\subsection*{Authors' Contributions}
CRediT taxonomy mapping:
\begin{itemize}
    \item \textbf{Aman Sharma}: Conceptualization, Methodology, Software, Formal Analysis and Investigation, Writing --- Original Draft, Writing --- Review \& Editing.
    \item \textbf{Manish Kumar}: Supervision, Resources, Writing --- Review \& Editing.
\end{itemize}

\subsection*{Acknowledgments}
We thank the Centre for Green Energy and Nanotechnology (C-GENT), Himachal Pradesh University, for providing computational infrastructure and institutional support throughout this research.

\end{document}
